Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-09-20 paper/main.tex
rewrite section without division

% --------------------------------------------------------------------------------------------------
\subsection{Without Division}
For now, we will focus on a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant position comparison scheme for our loss
function.
For this dataset, we assumed that all parameters which specify the interaction potential are
identical across all agents but growth rates \textbf{TODO check this} and the physical radii of the
cells are individual and we optimize them like any other global parameter.
We chose to optimize for 3 interaction parameters of the Mie potential $n,m,V_0$ and two parameters
which were individual for each cell, namely the radii $r_i$ and growth rates $\mu_i$.
We also included the damping constant $\lambda$ and removed any stochastic motion contributions
($D=0$).
This leads to a total of $4+2\times 6=16$ parameters.
To estimate these parameters we used a latin-hypercube~\cite{McKay1979} sweep across the parameter
space and performed a local minimization with the Nelder-Mead~\cite{Powell1973} method.

In figure~\ref{fig:parameter-estimates-single-step} we performed a sensitivity-analysis for every
parameter.
We can see that almost all parameters can be estimated by our optimization approach.
However, the damping constant (A) does not show any changes in the loss function when increasing its
value.
It hints at a non-identifiability~\cite{Raue2009} of the parameter $\lambda$.
But it also shows that the approach which we used is able to detect these problems.
As we continue to develop these methods, we expect to be able to reuse many of the existing
theoretical approaches such as likelihood estimation and uncertainty quantification which have
already been developed extensively for less complex systems such as \acp{ode}.
\textbf{TODO citations, also: is this discussion?}
Fortunately, \texttt{cellular\_raza} was chosen to construct our model such that we are flexible in
changing the underlying representation if should find that such unidentifiabilities are structural
(model-based) rather than practical (due to lack of data).
This allows us to iterate and find the model formulation which suits the data best but still retain
the mechanistic methodology of our approach.

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.49\textwidth]
        {figures/crm_fit/snapshots/predicted-001667.png}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.49\textwidth]
        {figures/crm_fit/celldiffs/diff-001667.png}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}%
    \caption{
        \textbf{TODO do we need this figure?}
        (A) A microscopic image with the predicted positions of the cells overlayed on top.
        (B) Differences between extracted and predicted positions.
        % TODO make this figure to scale
    }
    \label{fig:celldiffs-individual}
\end{figure}

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/damping.pdf}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/exponent-m.pdf}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/exponent-n.pdf}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/strength.pdf}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-0.pdf}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius.pdf}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
    \end{tikzonimage}
    \caption{
        Sensitivity analyses werer performed for every optimized parameter of our model.
        (A-F) show the curves of loss functions when varying this respective parameter and keeping
        all others constant at the value of the found optimum (blue dot).
        In the cases of (B-F), we can see that a minimum was achieved and thus the parameters could
        be estimated.
        However, for the case of the damping constant $\lambda$ in (A), we see that the upper bound
        is open.
        This suggests that our model could be reduced with regard to this parameter.
        We have exemplified the values for radii $r_i$ and growth rates $\mu_i$ with values of the
        first cell.
        The remaining plots can be found in the supplementary material in
        section~\ref{sec:supplement-parameter-profiles}.
    }
    \label{fig:parameter-estimates-single-step}
\end{figure}

>>>>

% --------------------------------------------------------------------------------------------------
\subsection{Without Division}
For now, we will focus on a time-interval which does not contain any division events and at which
the initial lag-phase of our bacteria os already passed.
By doing so, we can utilize the simpler and more performant position comparison scheme for our loss
function.
Furthermore, we need not to estimate the lag-time parameters which improves performance of our
optimization algorithm.
For this dataset, we assumed that all parameters which specify the interaction potential are
identical across all agents but growth rates of the cells are individual.

We chose to optimize for the damping of agents' movement within the medium, 3 interaction parameters
of the Mie potential $n,m,V_0$ a single radius for all agents and growth rates which were chosen
individually for each cell.
We removed any stochastic motion contributions by setting $D=0$.
This leads to a total of $5+1\times 6=11$ parameters.
To estimate these parameters we used the differential evolution~\cite{Storn1997} global optimization
algorithm which was initially seeded with a sample generated by the latin-hypercube~\cite{McKay1979}
method.
Finally, we performed a local minimization using the approach by Nelder-Mead~\cite{Powell1973}.

Figure~\ref{fig:parameter-estimates-single-step} shows profile-likelihoods for each optimized
parameter with respect to the optimization results described before.
We can see that all parameters can be estimated by our optimization approach and that no
non-identifiability~\cite{Raue2009} is present.

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.49\textwidth]
        {figures/crm_fit/snapshots/predicted-003889.png}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.49\textwidth]
        {figures/crm_fit/celldiffs/diff-003889.png}%
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}%
    \caption{
        (A) Final microscopic image of the time series with the predicted positions of the cells
        overlayed on top.
        (B) Differences between extracted and predicted positions.
        Note that the imaging-based comparison algorithm was not used to optimize this system but is
        chosen here as a tool to aid visualization of the prediction difference.
    }
\end{figure}

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/damping.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/exponent-m.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/exponent-n.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/strength.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-0.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
    \end{tikzonimage}
    \caption{
        We calculated the likelihood profiles for every optimized parameter of our model.
        All plots show how the value of the loss function varies when fixing a single parameter but
        still optimizing for the remaining ones.
        Every plot indicates that a minimum was achieved, thus every parameter could be identified.
        We show only one of the 6 individually-chosen growth rate as an example.
        The remaining plots show similar behaviour although with varying values and can be found in
        the supplementary material in section~\ref{sec:supplement-parameter-profiles}.
    }
    \label{fig:parameter-estimates-single-step}
\end{figure}
